---
title: '[R] Exploratory Visualization of Forest Fire Data'
author: "Erica V"
date: "19/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#Load the data and packages required
library(stats)
library(tidyverse)

forestfires <- read.csv("forestfires.csv")
colnames(forestfires)
```

## Introduction
The data is associated with a [scientific research paper](http://www3.dsi.uminho.pt/pcortez/fires.pdf) on predicting the occurencce of forest fires in Portugal using modeling techniques.

We know that the columns correspond to the following information:

* **X**: X-axis spatial coordinate within the Montesinho park map: 1 to 9 
* **Y**: Y-axis spatial coordinate within the Montesinho park map: 2 to 9 
* **month**: Month of the year: 'jan' to 'dec' 
* **day**: Day of the week: 'mon' to 'sun' 
* **FFMC**: Fine Fuel Moisture Code index from the FWI system: 18.7 to 96.20 
* **DMC**: Duff Moisture Code index from the FWI system: 1.1 to 291.3 
* **DC**: Drought Code index from the FWI system: 7.9 to 860.6 
* **ISI**: Initial Spread Index from the FWI system: 0.0 to 56.10 
* **temp**: Temperature in Celsius degrees: 2.2 to 33.30 
* **RH**: Relative humidity in percentage: 15.0 to 100 
* **wind**: Wind speed in km/h: 0.40 to 9.40 
* **rain**: Outside rain in mm/m2 : 0.0 to 6.4 
* **area**: The burned area of the forest (in ha): 0.00 to 1090.84 

A single row corresponds to the location of a fire and some characteristics about the fire itself. Higher water presence is typically associated with less fire spread, so we might expect the water-related variables (`DMC` and `rain`) to be related with `area`.

## Data Processing

```{r}
forestfires %>% pull(month) %>% unique
forestfires %>% pull(day) %>% unique

# organize the month and day of the year in order
month_order <- c("jan", "feb", "mar",
                 "apr", "may", "jun", 
                 "jul", "aug", "sep", 
                 "oct", "nov", "dec")

day_order <- c("sun", "mon", "tue", "wed", "thu", "fri", "sat")

forestfires <- forestfires %>%
  mutate(
    month = factor(month, levels = month_order),
    day = factor(day, levels = day_order)
  )
```

```{r}
# When do forest fires happen the most?

fires_month <- forestfires %>%
  group_by(month) %>%
  summarize(total_fires = n())

fires_month %>%
  ggplot(aes(x=month, y=total_fires)) +
  geom_col() +
  labs(
    title = "Number of Forest Fires by Month",
    y = "Fire Count",
    x = "Month"
  )+
  theme(
    panel.background = element_rect(fill = "white")
  )
```

We see a spike in forest fires in the months of August and September and other smaller spike in March.

```{r}
# Does day of the week play a role?
fires_day <- forestfires %>%
  group_by(day) %>%
  summarize(total_fires_day = n())

fires_day %>%
  ggplot(aes(x=day, y=total_fires_day)) +
  geom_col() +
  labs(
    title = "Number of Forest Fires by Day",
    y = "Fire Count",
    x = "Day of the Week"
  ) +
  theme(
    panel.background = element_rect(fill = "white")
  )
```
There seems to be an increase in forest fires on the weekend.

Let's investigate all other variables and see if any correlate with the same inclrease.

```{r}
# Plot all variables against time
forestfires_long <- forestfires %>%
  pivot_longer(
    cols = c("FFMC", "DMC", "DC", 
             "ISI", "temp", "RH", 
             "wind", "rain"),
    names_to = "column",
    values_to = "value"
  )

forestfires_long %>%
  ggplot(aes(x=month, y=value)) +
  geom_boxplot() +
  facet_wrap(vars(column), scale="free_y") +
  labs(
    title = "Variable changes over month",
    x = "Month",
    y = "Variable Value"
  )
```

From the plots we can see DC, DMC and temp also increase during August and September

Do these variables affect the forest fire area?

```{r}
forestfires_long %>%
  ggplot(aes(x=value, y=area)) +
  geom_point() +
  facet_wrap(vars(column), scale="free_x") +
  labs(
    title = "Relationship between other variables and area burned",
    x = "Value of Column",
    y = "Area Burned (hectare)"
    )
```

```{r}
# Remove outliers from data above to inspect better

forestfires_long %>% 
  filter(area < 300) %>% 
  ggplot(aes(x = value, y = area)) +
  geom_point() +
  facet_wrap(vars(column), scales = "free_x") +
  labs(
    title = "Relationships between other variables and area burned (area < 300)",
    x = "Value of column",
    y = "Area burned (hectare)"
  )
```

It appears that forest fire area covereage does not affect the severity of the forest fire.