---
title: 'Data Structures in R : Coronavirus Investigation'
author: "Erica"
date: "18/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Using an open-sourced [dataset](https://www.kaggle.com/lin0li/covid19testing), from Kaggle, concerning the number of Coronavirus (COVID-19) tests conducted over time. This dataset used here was collected between the 20th of January and the 1st of June 2020. Our analysis tries to provide an answer to this question: *Which countries have had the highest number of positive cases against the number of tests?*

```{r}
library(readr)
library(stats)

# Import the dataset
covid_df <- read.csv("covid19.csv")
```

```{r}
# Explore the data
# Show a global view of the dataset:
library(tibble)
glimpse(covid_df)

# Storing the column names in a variable and displaying them:
vector_cols <- colnames(covid_df)
vector_cols
```

The dataset contains `r ncol(covid_df)` columns and `r nrow(covid_df)` rows. 

This database provides information on the numbers (per day and cumulatively) of COVID-19 positive cases, deaths, tests performed and hospitalizations for each country through the column's names store in the variable `vector_cols`. 

```{r}
# Filter rows
library(dplyr)

# Filter the "All States" Province states and remove the `Province_State` column
covid_df_all_states <- covid_df %>% 
  filter(Province_State == "All States") %>% 
  select(-Province_State)

```

From the description of the dataset, we know the columns to be described as the following:

1. `Date`: Date
2. `Continent_Name`: Continent names
3. `Two_Letter_Country_Code`: Country codes
4. `Country_Region`: Country names
5. `Province_State`: States/province names; value is `All States` when state/provincial level data is not available
6. `positive`: Cumulative number of positive cases reported.
7. `active`: Number of actively cases on that **day**.
8. `hospitalized`: Cumulative number of hospitalized cases reported.
9. `hospitalizedCurr`: Number of actively hospitalized cases on that **day**.
10. `recovered`: Cumulative number of recovered cases reported.
11. `death`: Cumulative number of deaths reported.
12. `total_tested`: Cumulative number of tests conducted.
13. `daily_tested`: Number of tests conducted on the **day**; if daily data is unavailable, daily tested is averaged across number of days in between.
14. `daily_positive`: Number of positive cases reported on the **day**; if daily data is unavailable, daily positive is averaged across number of days in.

```{r}
# Filter columns
covid_df_all_states_daily <- covid_df_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)
```

```{r}
# We want the top ten cases countries. First we will group by country then aggregate the daily information.

covid_df_all_states_daily_sum <- covid_df_all_states_daily %>%
  group_by(Country_Region) %>%
  summarize(
    tested = sum(daily_tested),
    positive = sum(daily_positive),
    active = sum(active),
    hospitalized = sum(hospitalizedCurr)) %>%
  arrange(desc(tested))

covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_top_10
```

```{r}
# Extract vectors to do analysis
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized

# Name vectors
names(positive_cases) <- countries
names(tested_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries

# Identify the top three positive against tested cases
positive_tested = positive_cases / tested_cases
positive_tested_top_3 <- c("United Kingdom" = 0.11, "United States" = 0.10, "Turkey" = 0.08)
```

```{r}
# Consolidating data from the top three countries
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

# Creating a matrix of the top three country data
covid_mat <- rbind(united_kingdom, united_states, turkey)
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")
covid_mat
```

```{r}
# Summarize the results
question <-  "Which countries have had the highest number of positive cases against the number of tests?"

answer <- c("Positive tested cases" = positive_tested_top_3)

datasets <- list(
  original = covid_df,
  allstates = covid_df_all_states,
  daily = covid_df_all_states_daily,
  top_10 = covid_top_10
)
matrices <- list(covid_mat)
vectors <- list(vector_cols, countries)

data_structure_list <- list("dataframe" = datasets, "matrix" = matrices, "vector" = vectors)

covid_analysis_list <- list(question, answer, data_structure_list)
```

